Factoring RSA keys with Shor's algorithm

### 2020-06-03 paper/paper.tex
paper: Add screenshots+results, misc finishing touches

\documentclass[conference]{IEEEtran}
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{physics}
\begin{document}

@@

A mathematical description of RSA is out of scope for this paper, and can be found in \cite{rsa}. For the purposes of this paper, the cryptosystem can be thought of as a private key $A = \{p, q\}$, where $p$ and $q$ are primes known only to the receiver, and a public key $n=pq$, which can be disseminated. The public key allows any third party to encrypt messages, and the private key allows the receiver to decrypt messages intended for him.

\subsection{Reduction to the order-finding problem}
Miller \cite{miller} proved that the problem of factoring $N$ can be reduced to the "period-finding problem": for a given $a < N - 1$, find the smallest $r$ such that $a^r \equiv 1 \mod N$ (i.e. find the order of the algebraic group of integers coprime with $N$ and smaller than $N$, hence the name of "order-finding problem" by which it is found in literature). Specifically, depending on the choice of $a$ we have that either:
\begin{itemize}
\item $N$ and $a$ have some common prime factors, in which case $p=\gcd(N, a)$ returns one prime and $N/p$ returns the other;
\item let $r$ be the period defined above, then at least one of $p, q = \gcd(N, a^{r/2}\pm 1)$ are factors of $N$;
\item neither hold, we had a "bad" choice of $a$.
\end{itemize}

@@

The phase estimation algorithm can estimate $\theta=k/r$ to within a given error $\epsilon$: it returns $\widetilde\theta \in (\theta-\epsilon, \theta+\epsilon)$. We then need to minimize the error to a point where classical algorithms can recover $\theta$ from $\widetilde\theta$.

@@

Consider the family of unitary operators
\begin{equation}
U_a, U_{(a^2)}, U_{(a^4)}, U_{(a^8)}...
\end{equation}
All such operators $U_b$ are integer powers of $U_a$: it is easy to verify that if $b = a^n$ then $U_b = (U_a)^n$. For this reason they have the same eigenvectors; and despite being integer powers they are generally as easy/costly to implement as $U_a$, since one need only precompute $a^n \mod N$ by repeated squaring. Their eigenvalues, however, have far lower error than $U_a$: note that if $U_a$ has an eigenvalue $e^{i\phi}$, then $U_{(a^2)}$ has an eigenvalue $e^{2i\phi}$, $U_{(a^4)}$ has an eigenvalue $e^{4i\phi}$, and so on. The QPE will then estimate $\phi$, $2\phi$, $4\phi$... to within the same error $\epsilon$, thus reducing the error on $\phi$ to $\epsilon$, $\epsilon/2$, $\epsilon/4$...

@@

\subsection{RSA}
Although RSA keys are almost universally 1024, 2048 or 3072 bits long, such sizes are computationally infeasible, and practical concerns require us to use extremely small keys (around $N=30$ at most). The resulting circuits are otherwise too large and can take several minutes or hours to compute: empirically, we observe that a circuit for $N=33$ is executed in 9 minutes on the powerful IBM QASM simulator. Needless to say, a key with $N<30$ provides no actual security as it can be factored by hand, and only serves as a demonstration in place of larger keys.

Mainstream RSA implementations are designed with reasonable minimum key sizes: for instance, OpenSSL requires 384 bits or more, and will fail with an error for shorter keys. This requirement partly stems from the use of padding schemes, which add desirable security properties such as non-determinism (the same message encrypted several times will have several different encryptions; this is not true with plain RSA), preventing partial decryption, and others. As we only want to explore the feasibility of factoring the RSA key, such security properties are not of interest to us, and we can use a custom RSA implementation with tiny keys and no padding schemes.

@@

We present a dramatic improvement upon this algorithm that brings the time spent constructing the circuit from 127 seconds (Qiskit's implementation at the time of writing) to 6 for $N=33$. The resulting circuit is identical and thus has the same execution time; this improvement merely makes the circuit construction faster.

Simple line profiling shows that the Qiskit implementation spends a large amount of time "composing" subcircuits together (for instance, appending an IQFT circuit as part of the QPE routine). Delving into the library code shows that the composition of circuits is implemented by transforming the operands into DAGs\footnote{DAG: Directed Acyclic Graph.}, merging those, and then retransforming back into a circuit. We report an abridged version:

@@

Further profiling shows that the conversion from and to DAGs is very computationally intensive with respect to the other steps. The key intuition is that multiple compositions are therefore wasteful, as they spend a lot of time converting intermediate values back and forth; representing them as DAGs instead, only converting to and from DAGs respectively at the beginning and at the end of the algorithm, allows for much more efficient composition. In reality the composition takes additional arguments (most importantly the qubit indices on which to compose), which need to be stored along with the DAG - we call this object a "tagged DAG". This simple modification then allows to construct the circuit correctly with dramatic speed improvements.

This improvement was discovered independently by Thomas Metcalfe, who submitted it in a slightly different form on May 25 as part of pull request \#975 to Qiskit Aqua. As the PR is not yet merged at the time of writing, we have released our implementation \cite{github-repo} as a separate repository.

@@

%\subsection{\LaTeX-Specific Advice}

%Please don't use the \verb|{eqnarray}| equation environment. Use \verb|{align}| or \verb|{IEEEeqnarray}| instead. The \verb|{eqnarray}|environment leaves unsightly spaces around relation symbols.

%\subsection{Identify the Headings}
%Headings, or heads, are organizational devices that guide the reader through 
%your paper. There are two types: component heads and text heads.

%Component heads identify the different components of your paper and are not 
%topically subordinate to each other. Examples include Acknowledgments and 
%References and, for these, the correct style to use is ``Heading 5''. Use 
%``figure caption'' for your Figure captions, and ``table head'' for your 
%table title. Run-in heads, such as ``Abstract'', will require you to apply a 
%style (in this case, italic) in addition to the style provided by the drop 
%down menu to differentiate the head from the text.

%Text heads organize the topics on a relational, hierarchical basis. For 
%example, the paper title is the primary text head because all subsequent 
%material relates and elaborates on this one topic. If there are two or more 
%sub-topics, the next level head (uppercase Roman numerals) should be used 
%and, conversely, if there are not at least two sub-topics, then no subheads 
%should be introduced.

%\subsection{Figures and Tables}
%\paragraph{Positioning Figures and Tables} Place figures and tables at the top and 
%bottom of columns. Avoid placing them in the middle of columns. Large 
%figures and tables may span across both columns. Figure captions should be 
%below the figures; table heads should appear above the tables. Insert 
%figures and tables after they are cited in the text. Use the abbreviation 
%``Fig.~\ref{fig}'', even at the beginning of a sentence.
%
%\begin{table}[htbp]
%\caption{Table Type Styles}
%\begin{center}
%\begin{tabular}{|c|c|c|c|}
%\hline
%\textbf{Table}&\multicolumn{3}{|c|}{\textbf{Table Column Head}} \\
%\cline{2-4} 
%\textbf{Head} & \textbf{\textit{Table column subhead}}& \textbf{\textit{Subhead}}& \textbf{\textit{Subhead}} \\
%\hline
%copy& More table copy$^{\mathrm{a}}$& &  \\
%\hline
%\multicolumn{4}{l}{$^{\mathrm{a}}$Sample of a Table footnote.}
%\end{tabular}
%\label{tab1}
%\end{center}
%\end{table}
%
%\begin{figure}[htbp]
%\centerline{\includegraphics{fig1.png}}
%\caption{Example of a figure caption.}
%\label{fig}
%\end{figure}

%Figure Labels: Use 8 point Times New Roman for Figure labels. Use words 
%rather than symbols or abbreviations when writing Figure axis labels to 
%avoid confusing the reader. As an example, write the quantity 
%``Magnetization'', or ``Magnetization, M'', not just ``M''. If including 
%units in the label, present them within parentheses. Do not label axes only 
%with units. In the example, write ``Magnetization (A/m)'' or ``Magnetization 
%\{A[m(1)]\}'', not just ``A/m''. Do not label axes with a ratio of 
%quantities and units. For example, write ``Temperature (K)'', not 
%``Temperature/K''.

\begin{thebibliography}{00}
\bibitem{shor} P. W. Shor, "Algorithms for quantum computation: discrete logarithms and factoring", SIAM J.Sci.Statist.Comput. 26
\bibitem{rsa} R. Rivest, A. Shamir, L. Adleman, "A Method for Obtaining Digital Signatures and Public-Key Cryptosystems", Communications of the ACM 21
\bibitem{miller} G. L. Miller, "Riemann's hypothesis and tests for primality", J. Comput. System Sci.
\bibitem{vbe} V. Vedral, A. Barenco, A. Ekert, "Quantum Networks for Elementary Arithmetic Operations", Physical Review A 54
\bibitem{toffoli} D. Maslov, G. W. Dueck, D. M. Miller, "Toffoli network synthesis with templates," IEEE Transactions on Computer-Aided Design of Integrated Circuits and Systems
\bibitem{ibm} IBM, "Shor's algorithm",\\https://quantum-computing.ibm.com/docs/guide/q-algos/shor-s-algorithm
\bibitem{github-repo} G. Muscarello, https://github.com/CapacitorSet/qiskit-fast-shor
\end{thebibliography}
\vspace{12pt}

>>>>

\documentclass[conference]{IEEEtran}
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{physics}
\usepackage{caption}
\usepackage{stfloats}

\begin{document}

@@

A mathematical description of RSA is out of scope for this paper, and can be found in \cite{rsa}. For the purposes of this paper, the cryptosystem can be thought of as a private key $\{p, q\}$, where $p$ and $q$ are primes known only to the receiver, and a public key $N=pq$, which can be disseminated. The public key allows any third party to encrypt messages, and the private key allows the receiver to decrypt messages intended for him.

\subsection{Reduction to the order-finding problem}
Miller \cite{miller} proved that the problem of factoring $N$ can be reduced to the "period-finding problem": for a given $a < N - 1$, find the smallest $r$ such that $a^r \equiv 1 \mod N$ (i.e. find the order of the algebraic group of integers coprime with $N$ and smaller than $N$, hence the name of "order-finding problem" by which it is found in literature). Specifically, depending on the choice of $a$ we have that either:
\begin{itemize}
\item $N$ and $a$ have a common prime factor, in which case $p=\gcd(N, a)$ returns one prime and $N/p$ returns the other;
\item let $r$ be the period defined above, then at least one of $p, q = \gcd(N, a^{r/2}\pm 1)$ are factors of $N$;
\item neither hold, we had a "bad" choice of $a$.
\end{itemize}

@@

The phase estimation algorithm can estimate $\theta=k/r$ to within a given error $\epsilon$: it returns $\widetilde\theta \in (\theta-\epsilon, \theta+\epsilon)$. We then need to minimize the error to a point where classical algorithms (in this case, continued fraction expansion) can recover $\theta$ from $\widetilde\theta$.

@@

Consider the family of unitary operators
\begin{equation}
U_a, U_{(a^2)}, U_{(a^4)}, U_{(a^8)}...
\end{equation}
All such operators $U_b$ are integer powers of $U_a$: it is easy to verify that if $b = a^n$ then $U_b = (U_a)^n$. For this reason they have the same eigenvectors, but despite being integer powers they are only as hard to implement as $U_a$, since one need only precompute $a^n \mod N$ by repeated squaring. Their eigenvalues, however, have far lower error than $U_a$: note that if $U_a$ has an eigenvalue $e^{i\phi}$, then $U_{(a^2)}$ has an eigenvalue $e^{2i\phi}$, $U_{(a^4)}$ has an eigenvalue $e^{4i\phi}$, and so on. The QPE will then estimate $\phi$, $2\phi$, $4\phi$... to within the same error $\epsilon$, thus reducing the error on $\phi$ to $\epsilon$, $\epsilon/2$, $\epsilon/4$...

@@

\subsection{RSA}
RSA keys are typically 1024 or 2048 bits long, with some being 3072 or 4096 bits long. However, factoring such large keys with Shor's algorithm is computationally infeasible, and practical concerns require us to use extremely small keys (around $N=30$ at most). The resulting circuits are otherwise too large and can take several minutes or hours to compute: empirically, we observe that a circuit for $N=33$ is executed in 10 minutes on the powerful IBM QASM simulator. Needless to say, such small keys provide no actual security as they can be factored by hand, and only serve as a demonstration in place of larger keys.

Mainstream RSA implementations are designed with reasonable minimum key sizes: for instance, OpenSSL requires 384 bits or more, and will fail with an error for shorter keys. This requirement partly stems from the use of padding schemes, which add desirable security properties such as non-determinism (the same message encrypted several times will have several different encryptions; this is not true with plain RSA), preventing partial decryption, and others. As we only want to explore the feasibility of factoring the RSA key, such security properties are not of interest to us, and we can use a custom RSA implementation with tiny keys and no padding scheme.

@@

We present a dramatic improvement upon this algorithm that brings the time spent constructing the circuit from 127 seconds (using Qiskit's implementation at the time of writing) to 6 seconds for $N=33$. The resulting circuit is identical and thus has the same execution time; this improvement merely makes the circuit construction faster.

Simple line profiling shows that the Qiskit implementation spends a large amount of time composing subcircuits together (for instance, appending an IQFT circuit as part of the QPE routine). Delving into the library code shows that the composition of circuits is implemented by transforming the operands into DAGs\footnote{DAG: Directed Acyclic Graph.}, merging those, and then retransforming back into a circuit. We report an abridged version of the method:

@@

Further profiling shows that the conversion from and to DAGs is very computationally intensive with respect to the other steps. The key intuition is that sequential compositions are therefore wasteful, as they spend a lot of time converting intermediate values back and forth; representing them as DAGs instead, only converting to and from DAGs respectively at the beginning and at the end of the algorithm, allows for much more efficient composition. In reality the composition takes additional arguments (most importantly the qubit indices on which to compose), which need to be stored along with the DAG - we call this object a "tagged DAG". This simple modification then allows to construct the circuit correctly with dramatic speed improvements.

This improvement was discovered independently by Thomas Metcalfe, who submitted it in a slightly different form on May 25 as part of pull request \#975 to Qiskit Aqua. As the PR is not yet merged at the time of writing, we have released our implementation \cite{github-repo-shor} as a separate repository.

@@

We have released the code for the demonstration, together with this paper, as a public Git repository \cite{github-repo-main}.

\section{Results and conclusions}
We demonstrate that Shor's algorithm can be used to factor RSA keys and compromise encrypted channels. We present three cases:
\begin{itemize}
\item $N=15$, factored on a local simulator (Aer backend) in a few seconds;
\item $N=21$, factored on the IBM QASM simulator in 40 seconds;
\item $N=33$, which gave inconclusive results after approximately 10 minutes on the IBM QASM simulator.
\end{itemize}

We notice that with the increase of $N$ the circuit depth and the execution times rise greatly, making local simulation impractical after a certain point. However, even powerful simulators like the one provided by IBM struggle to factor even modest values of $N$, which in turn are extremely tiny in comparison to keys used in the real world (the smaller ones being 1024 bit long).

While performance issues could in theory be addressed with real quantum computers with sufficiently many qubits, we also observe that the demonstration often fails for $N=33$. We attribute this to the large amount of noise and decoherence accumulated in the computation.

Overall, we find that Shor's algorithm is a simple and well-studied method to solve the factoring problem, but at the current time it is vastly unfeasible as a tool against real RSA keys due to the limits of current quantum computing technology.

\begin{figure*}[b]
   \centering
   \includegraphics[width=\textwidth,height=\textheight,keepaspectratio]{2020-06-03_20-38.png}
   \caption{A demonstration of the attack on the public key $e=3$, $N=15$. The algorithm was executed on a local simulator (Aer backend); the timestamps show that the factorization was carried out in a few seconds.\\Top left: the client interface. Top right: the server interface. Bottom: the attacker interface.}
\end{figure*}

\begin{figure*}[t]
   \centering
   \includegraphics[width=\textwidth,height=\textheight,keepaspectratio]{2020-06-03_23-29.png}
   \caption{A demonstration of the attack on the public key $e=3$, $N=21$. The algorithm was executed on the IBM QASM simulator.}
\end{figure*}

\begin{figure*}[t]
   \centering
   \includegraphics[width=\textwidth,height=\textheight,keepaspectratio]{2020-06-03_20-54.png}
   \caption{A demonstration of the attack on the public key $e=3$, $N=33$. The algorithm was executed on the IBM QASM simulator, and gave inconclusive results.}
\end{figure*}

\begin{figure*}[t]
   \centering
   \includegraphics[width=\textwidth,height=\textheight,keepaspectratio]{2020-06-03_20-56.png}
   \caption{Timings for factorizing $pq=33$ on the IBM QASM simulator.}
\end{figure*}

\begin{thebibliography}{00}
\bibitem{shor} P. W. Shor, "Algorithms for quantum computation: discrete logarithms and factoring", SIAM J.Sci.Statist.Comput. 26
\bibitem{rsa} R. Rivest, A. Shamir, L. Adleman, "A Method for Obtaining Digital Signatures and Public-Key Cryptosystems", Communications of the ACM 21
\bibitem{miller} G. L. Miller, "Riemann's hypothesis and tests for primality", J. Comput. System Sci.
\bibitem{vbe} V. Vedral, A. Barenco, A. Ekert, "Quantum Networks for Elementary Arithmetic Operations", Physical Review A 54
\bibitem{toffoli} D. Maslov, G. W. Dueck, D. M. Miller, "Toffoli network synthesis with templates," IEEE Transactions on Computer-Aided Design of Integrated Circuits and Systems
\bibitem{ibm} IBM, "Shor's algorithm",\\https://quantum-computing.ibm.com/docs/guide/q-algos/shor-s-algorithm
\bibitem{github-repo-main} G. Muscarello, https://github.com/CapacitorSet/shor-demonstration
\bibitem{github-repo-shor} G. Muscarello, https://github.com/CapacitorSet/qiskit-fast-shor
\end{thebibliography}
\vspace{12pt}
